---
title: "Reproducible Research Course Project 1 - Activity Monitoring Data"
output: html_document
---

## Part 1: Code for reading in the dataset and/or processing the data
```{r, echo=TRUE}
setwd('C:/Users/mtsc7/Desktop/coursera')
activity <- read.csv("activity.csv")
summary(activity)
str(activity)
```


## Part 2: Histogram of the total number of steps taken each day
```{r, echo=TRUE}
steps_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(steps_day$steps, xlab = "Steps each day", main = "Total number of steps taken each day", col = "blue")
```


## Part 3: Mean and median number of steps taken each day
```{r, echo=TRUE}
mean_steps <- mean(steps_day$steps)
mean_steps <- round(mean_steps, digits = 1)
print(mean_steps)

median_steps <- median(steps_day$steps)
median_steps <- round(median_steps, digits = 1)
print(median_steps)
```
ANSWER: The mean step is `r mean_steps` and median step is `r median_steps`.


##Part 4: Time series plot of the average number of steps taken
```{r, echo=TRUE}
mean_steps_ts <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm = TRUE)
plot(mean_steps_ts$interval, mean_steps_ts$steps, type = "l", col = "blue", xlab = "Intervals", ylab = "Total steps per interval", main = "Number of steps per interval (averaged) (Exclude NA)")
```


##Part 5: The 5-minute interval that, on average, contains the maximum number of steps
```{r, echo=TRUE}
max_steps <-max(mean_steps_ts$steps)
max_steps_interval <- mean_steps_ts$interval[which(mean_steps_ts$steps == max_steps)]
print(max_steps_interval)

max_steps <- round(max_steps, digits = 1)
print(max_steps)
```
ANSWER: The maximum steps taken is `r max_steps` and occurred at interval `r max_steps_interval`.

## Part 6: Code to describe and show a strategy for imputing missing data
### STRATEGY - The missing data will be imputed with the mean number of steps across all days with non-NA data for that particular interval.
```{r, echo=TRUE}
total_na <- sum(is.na(activity))

Steps_Interval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
activity_split <- split(activity, activity$interval)

#Impute missing data for each interval
for(i in 1:length(activity_split)){
  activity_split[[i]]$steps[is.na(activity_split[[i]]$steps)] <- Steps_Interval[i]
}
activity_imputed <- do.call("rbind", activity_split)
activity_imputed <- activity_imputed[order(activity_imputed$date) ,]
```
ANSWER: The total number of missing value is `r total_na`.


##Part 7: Histogram of the total number of steps taken each day after missing values are imputed
```{r, echo=TRUE}
steps_imputed <- tapply(activity_imputed$steps, activity_imputed$date, sum)
hist(steps_imputed, xlab = "Number of Steps", main = "Number of steps per Day (After Imputed)", col = "blue")

#Mean and median number of steps taken each day
imputed_mean_steps <- mean(steps_imputed)
imputed_mean_steps <- round(imputed_mean_steps, digits = 1)

imputed_median_steps <- median(steps_imputed)
imputed_median_steps <- round(imputed_median_steps, digits = 1)

#Compare the results in a dataframe
compare_mean_median <- data.frame(c(mean_steps, median_steps), c(imputed_mean_steps, imputed_median_steps))
colnames(compare_mean_median) <- c("Before Impute NA", "After impute NA")
rownames(compare_mean_median) <- c("mean", "median")

library(xtable)
xt <- xtable(compare_mean_median)
print(xt, type  = "html")
```
#### ANSWER - Imputing missing values does not change the mean value (`r imputed_mean_steps`) and insignificantly increased the median value (from `r median_steps` to `r imputed_median_steps`).


## Step 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r, echo=TRUE}
activity_imputed$wktype <- ifelse(weekdays(as.Date(activity_imputed$date)) == "Saturday" | weekdays(as.Date(activity_imputed$date)) == "Sunday", "weekend", "weekday")

#Average steps per interval for weekdays and weekends
steps_weekday <- tapply(activity_imputed[activity_imputed$wktype == "weekday" ,]$steps, activity_imputed[activity_imputed$wktype == "weekday" ,]$interval, mean, na.rm = TRUE)
steps_weekend <- tapply(activity_imputed[activity_imputed$wktype == "weekend" ,]$steps, activity_imputed[activity_imputed$wktype == "weekend" ,]$interval, mean, na.rm = TRUE)

#Create a 2 panel plot for weekday and weekend
par(mfrow=c(1,2))

plot(as.numeric(names(steps_weekday)), 
     steps_weekday, 
     xlab = "Interval", 
     ylab = "Steps", 
     main = "Activity in Weekdays", 
     type = "l")

plot(as.numeric(names(steps_weekend)), 
     steps_weekend, 
     xlab = "Interval", 
     ylab = "Steps", 
     main = "Activity in Weekends", 
     type = "l")
```

#### ANSWER: Activity (by number of steps) have started earlier in the weekdays and lasted over a shorter duration (by the occurance of  high peaks). However activity on weekends started slightly later and spread over longer duration.